\section{Conclusion}
\label{sec:conclusion}

We presented the Neural Residual ODE framework for ENSO forecasting, systematically investigating how to combine physics-based structure with machine learning flexibility for climate prediction under severe data limitations. Our model, NXRO, and its 43 model variants, are evaluated on rigorous out-of-sample forecasting and yield insights relevant to both the climate science and machine learning communities. The central result of this work is that simple hybrid architectures outperform both the full physics-based XRO model and pure nonlinear deep learning approaches when the observational dataset is scarce, and adding synthetic climate simulation data can introduce systematic biases that hamper the model performance. This result carries implications for practitioners: modern deep learning architectures do not automatically transfer to small-data scientific domains, and the choice of synthetic datasets matter. The hybrid approach, which encodes well-understood physics as architectural structure while learning corrections from data, emerges as the optimal strategy for the small-data regime characteristic of climate science and many other scientific domains. In practice, it points to a strategy to construct a physically meaningful and accurate model without relying on massive scale simulations. % We hope to extend this framework to application to other scientific domains, such as fluid and molecular dynamics, and other climate indices. 


\section{Limitations and Ethical Considerations}

Our results speak to ongoing debates about the role of physics in AI4science and demonstrates that for the task of ENSO forecasting, a physics-inspired simple neural model can outperform fully nonlinear neural models, even with additional simulation data. Nevertheless, our study is limited to ENSO forecast, and we hope to extend it to other scientific domains, such as fluid and molecular dynamics, and other climate indices. In addition, since the choice of synthetic datasets matter, we hope to more thoroughly examine other climate simulation datasets as well. For the ethical implications, we are not aware of any issues related to data privacy, consent, bias, and potential misuses. Our dataset will be open sourced, consisting of observation data and climate model simulations, none of which has ethical implications. 